---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

options(tidyverse.quiet = TRUE)
```
# leapfrog

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/mrc-ide/leapfrog.svg?branch=master)](https://travis-ci.org/mrc-ide/leapfrog)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/mrc-ide/leapfrog?branch=master&svg=true)](https://ci.appveyor.com/project/mrc-ide/leapfrog)
[![Codecov test coverage](https://codecov.io/gh/mrc-ide/leapfrog/branch/master/graph/badge.svg)](https://codecov.io/gh/mrc-ide/leapfrog?branch=master)
<!-- badges: end -->

Leapfrog is a multistate population projection model for demographic and HIV epidemic estimation.

The name _leapfrog_ is in honor of [Professor](https://blogs.lshtm.ac.uk/alumni/2018/07/16/obituary-professor-basia-zaba/) Basia [Zaba](https://translate.google.co.uk/#view=home&op=translate&sl=pl&tl=en&text=%C5%BBaba).

## Installation

Install the development version from [GitHub](https://github.com/mrc-ide/leapfrog) via devtools:

``` r
# install.packages("devtools")
devtools::install_github("mrc-ide/leapfrog")
```

## Example

Construct a sparse Leslie matrix:

```{r example}
library(tidyverse)
library(leapfrog)
library(popReconstruct)

data(burkina_faso_females)

make_leslie_matrixR(sx = burkina.faso.females$survival.proportions[,1],
                    fx = burkina.faso.females$fertility.rates[4:10, 1],
                    srb = 1.05,
                    interval = 5,
                    fx_idx = 4)
```

Simulate a cohort component population projection:

```{r ccmpp_sim}
pop_proj <- ccmppR(basepop = as.numeric(burkina.faso.females$baseline.pop.counts),
                   sx = burkina.faso.females$survival.proportions,
                   fx = burkina.faso.females$fertility.rates[4:10, ],
                   gx = burkina.faso.females$migration.proportions,
                   srb = rep(1.05, ncol(burkina.faso.females$survival.proportions)),
                   interval = 5,
                   fx_idx = 4)
pop_proj$population[ , c(1, 2, 10)]

```


### TMB 

Calculate a population projection in TMB. Carry forward 2000 values for two further periods
to explore projections.

```{r tmb_example}

basepop_init <- as.numeric(burkina.faso.females$baseline.pop.counts)

sx_init <- burkina.faso.females$survival.proportions
sx_init <- cbind(sx_init, `2005` = sx_init[ , "2000"], `2010` = sx_init[ , "2000"])

fx_init <- burkina.faso.females$fertility.rates[4:10, ]
fx_init <- cbind(fx_init, `2005` = fx_init[ , "2000"], `2010` = fx_init[ , "2000"])

gx_init <- burkina.faso.females$migration.proportions
gx_init <- cbind(gx_init, `2005` = gx_init[ , "2000"], `2010` = gx_init[ , "2000"])

log_basepop_mean <- as.vector(log(basepop_init))
logit_sx_mean <- as.vector(qlogis(sx_init))
log_fx_mean <- as.vector(log(fx_init))
gx_mean <- as.vector(gx_init)
  
data <- list(log_basepop_mean = log_basepop_mean,
             logit_sx_mean = logit_sx_mean,
             log_fx_mean = log_fx_mean,
             gx_mean = gx_mean,
             srb = rep(1.05, ncol(sx_init)),
             interval = 5,
             n_steps = ncol(sx_init),
             fx_idx = 4L,
             n_fx = 7L,
             census_log_pop = log(burkina.faso.females$census.pop.counts),
             census_year_idx = c(4L, 6L, 8L, 10L))
par <- list(log_tau2_logpop = 0,
            log_tau2_sx = 0,
            log_tau2_fx = 0,
            log_tau2_gx = 0,
            log_basepop = log_basepop_mean,
            logit_sx = logit_sx_mean,
            log_fx = log_fx_mean,
            gx = gx_mean)

obj <- make_tmb_obj(data, par)

init_sim <- obj$report()
matrix(init_sim$population, nrow = 17)[ , data$census_year_idx]

obj$fn()


input <- list(data = data, par_init = par)
fit <- fit_tmb(input)

fit[1:6]

```

Sample from posterior distribution and generate outputs

```{r sample_tmb}
fit <- sample_tmb(fit)



colnames(fit$sample$population) <- 1:ncol(fit$sample$population)
colnames(fit$sample$migrations) <- 1:ncol(fit$sample$migrations)
colnames(fit$sample$fx) <- 1:ncol(fit$sample$fx)

init_pop_mat <- ccmpp_leslieR(basepop_init, sx_init, fx_init, gx_init,
                              srb = rep(1.05, ncol(sx_init)),
                              interval = 5, fx_idx = 4)

df <- crossing(year = seq(1960, 2015, 5),
               sex = "female",
               age_group = c(sprintf("%02d-%02d", 0:15*5, 0:15*5+4), "80+")) %>%
  mutate(init_pop = as.vector(init_pop_mat))

census_pop <- crossing(sex = "female",
                       age_group = c(sprintf("%02d-%02d", 0:15*5, 0:15*5+4), "80+")) %>%
  bind_cols(as_tibble(burkina.faso.females$census.pop.counts)) %>%
  gather(year, census_pop, `1975`:`2005`) %>%
  type_convert(cols(year = col_double()))

df <- df %>%
  left_join(census_pop) %>%
  bind_cols(as_tibble(fit$sample$population)) %>%
  gather(sample, value, `1`:last_col())

agepop <- df %>%
  group_by(year, sex, age_group) %>%
  summarise(init_pop = mean(init_pop),
            census_pop = mean(census_pop),
            mean = mean(value),
            lower = quantile(value, 0.025),
            upper = quantile(value, 0.975))

totalpop <- df %>%
  group_by(year, sample) %>%
  summarise(init_pop = sum(init_pop),
            census_pop = sum(census_pop),
            value = sum(value)) %>%
  group_by(year) %>%
  summarise(init_pop = mean(init_pop),
            census_pop = mean(census_pop),
            mean = mean(value),
            lower = quantile(value, 0.025),
            upper = quantile(value, 0.975))
```

```{r, fig.height = 4, fig.width = 5, fig.align = "center", out.width = "60%"}
ggplot(totalpop, aes(year, mean, ymin = lower, ymax = upper)) +
  geom_ribbon(alpha = 0.2) +
  geom_line() +
  geom_line(aes(y = init_pop), linetype = "dashed") +
  geom_point(aes(y = census_pop), shape = 4, color = "darkred", stroke = 2) +
  scale_y_continuous("Total population (millions)", labels = scales::number_format(scale = 1e-6)) +
  expand_limits(y = 0) +
  labs(x = NULL) +
  theme_light() +
  ggtitle("BFA females: total population")
```

```{r, fig.height = 5, fig.width = 7, fig.align = "center", out.width = "80%"}
ggplot(agepop, aes(age_group, mean, ymin = lower, ymax = upper, group = 1)) +
  geom_ribbon(alpha = 0.2) +
  geom_line() +
  geom_line(aes(y = init_pop), linetype = "dashed") +
  geom_point(aes(y = census_pop), color = "darkred") +
  facet_wrap(~year, scales = "free_y") + 
  scale_y_continuous("Total population (thousands)", labels = scales::number_format(scale = 1e-3)) +
  expand_limits(y = 0) +
  theme_light() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid = element_blank()) +
  ggtitle("BFA females: population by age")
```

Posterior distribution for TFR:

```{r asfr, fig.height = 4, fig.width = 5, fig.align = "center", out.width = "60%"}

asfr <- crossing(year = seq(1960, 2010, 5),
                 age_group = sprintf("%02d-%02d", 3:9*5, 3:9*5+4)) %>%
  mutate(init_asfr = as.vector(fx_init)) %>%
  bind_cols(as_tibble(fit$sample$fx)) %>%
  gather(sample, value, `1`:last_col()) 

tfr <- asfr %>%
  group_by(year, sample) %>%
  summarise(init_tfr = 5 * sum(init_asfr),
            value = 5 * sum(value)) %>%
  group_by(year) %>%
  summarise(init_tfr = mean(init_tfr),
         mean = mean(value),
         lower = quantile(value, 0.025),
         upper = quantile(value, 0.975))

ggplot(tfr, aes(year, mean, ymin = lower, ymax = upper)) +
  geom_ribbon(alpha = 0.2) +
  geom_line() +
  geom_line(aes(y = init_tfr), linetype = "dashed") +
  scale_y_continuous("Total fertility rate", limits = c(5, 9)) +
  labs(x = NULL) +
  theme_light() +
  theme(panel.grid = element_blank()) +
  ggtitle("BFA: total fertility rate")

```

```{r migrations, fig.height = 4, fig.width = 5, fig.align = "center", out.width = "60%"}

migrations <- crossing(year = seq(1960, 2010, 5),
                       age_lower = 0:16*5) %>%
  mutate(init_migrations = init_sim$migrations) %>%
  bind_cols(as_tibble(fit$sample$migrations)) %>%
  gather(sample, value, `1`:last_col()) 

total_migrations <- migrations %>%
  group_by(year, sample) %>%
  summarise(init_migrations = 5 * sum(init_migrations),
            value = 5 * sum(value)) %>%
  group_by(year) %>%
  summarise(init_migrations = mean(init_migrations),
         mean = mean(value),
         lower = quantile(value, 0.025),
         upper = quantile(value, 0.975))

ggplot(total_migrations, aes(year, mean, ymin = lower, ymax = upper)) +
  geom_ribbon(alpha = 0.2) +
  geom_line() +
  geom_line(aes(y = init_migrations), linetype = "dashed") +
  scale_y_continuous("Net migration (1000s)",
                     labels = scales::number_format(scale=1e-3)) +
  labs(x = NULL) +
  theme_light() +
  theme(panel.grid = element_blank()) +
  ggtitle("BFA: total net migrations (1000s)")

```


## Code design

### Simulation model

The simulation model is implemented as templated C++ code in `src/ccmpp.h`.
This is  the simulation model may be developed as a standalone C++ library
that can be called by other software without requiring R-specific code features.
The code uses header-only open source libraries to maximize portability.


### Statistical inference

The objective function (the negative log posterior density) is implemented 
in templated C++ code in the script `src/TMB/ccmpp_tmb.hpp` using probability 
functions from the Template Model Builder ([TMB](https://github.com/kaskr/adcomp/wiki)) 
R package. The TMB package provides estimates of the gradient
of the objective function via automatic differentiation and Laplace approximation
to integrate random effects. 

The posterior density can also be sampled from using Hamiltonian Monte Carlo in 
Stan using the [_tmbstan_](https://cran.r-project.org/web/packages/tmbstan/index.html) package.

Latent Gaussian models for model components (fertiliy rates, mortality rates,
migration rates) are implemented using linear algebra tools from the _Eigen_ 
package. Predicted values are coerced into arrays for the CCMPP model inputs.


### R functions

The file `src/ccmppR.cpp` contains R wrapper functions for the model simulation 
via [Rcpp](http://dirk.eddelbuettel.com/code/rcpp.html) and 
[RcppEigen](http://dirk.eddelbuettel.com/code/rcpp.eigen.html).


## Development notes

### Simulation model

* The CCMPP model is implemented as a sparse Leslie matrix formulation and using
  direct calculation of the projection in arrays so that interim outputs (deaths,
  births, migrations) are also saved. The array-based implementation appears to 
  be faster.
* Class structure for popualtion projection model needs review.
* Specifying static dimensions for the state space may improve efficiency. This 
  should be possible for common options (5x5 year, 1x1 year) through templating.
* The model was implemented using _Eigen_ containers following the initial sparse
  Leslie matrix specification. However, multi-dimensional arrays in the _boost_
  library may be preferable. 

### TMB

* Further investigation is needed about the portability of AD objective function
  DLLs outside of the R environment.

TMB model code and testing are implemented following templates from the 
[`TMBtools`](https://github.com/mlysy/TMBtools) package with guidance for package 
development with both TMB models and Rcpp code. 
  * To add a new TMB model, save the model template in the `src/TMB` with extension `.hpp`. The model name must match the file name. The signature is slightly different -- see other `.hpp` files for example.
  * Call `TMBtools::export_models()` to export the new TMB model to the meta-model list.
  * When constructing the objective function with `TMB::MakeADFun`, use `DLL= "leapfrog_TMBExports"` and add an additional value `model = "<model_name>"` to the `data` list.
