---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

options(tidyverse.quiet = TRUE)
```
# leapfrog

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/mrc-ide/leapfrog.svg?branch=master)](https://travis-ci.org/mrc-ide/leapfrog)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/mrc-ide/leapfrog?branch=master&svg=true)](https://ci.appveyor.com/project/mrc-ide/leapfrog)
[![Codecov test coverage](https://codecov.io/gh/mrc-ide/leapfrog/branch/master/graph/badge.svg)](https://codecov.io/gh/mrc-ide/leapfrog?branch=master)
<!-- badges: end -->

Leapfrog is a multistate population projection model for demographic and HIV epidemic estimation.

The name _leapfrog_ is in honor of [Professor](https://blogs.lshtm.ac.uk/alumni/2018/07/16/obituary-professor-basia-zaba/) Basia [Zaba](https://translate.google.co.uk/#view=home&op=translate&sl=pl&tl=en&text=%C5%BBaba).

## Installation

Install the development version from [GitHub](https://github.com/mrc-ide/leapfrog) via devtools:

``` r
# install.packages("devtools")
devtools::install_github("mrc-ide/leapfrog")
```

## Example

Construct a sparse Leslie matrix:

```{r example}
library(tidyverse)
library(leapfrog)
library(popReconstruct)

data(burkina_faso_females)

make_leslie_matrixR(sx = burkina.faso.females$survival.proportions[,1],
                    fx = burkina.faso.females$fertility.rates[4:10, 1],
                    srb = 1.05,
                    age_span = 5,
                    fx_idx = 4)
```

Simulate a cohort component population projection:

```{r ccmpp_sim}
pop_proj <- ccmppR(basepop = as.numeric(burkina.faso.females$baseline.pop.counts),
                   sx = burkina.faso.females$survival.proportions,
                   fx = burkina.faso.females$fertility.rates[4:10, ],
                   gx = burkina.faso.females$migration.proportions,
                   srb = rep(1.05, ncol(burkina.faso.females$survival.proportions)),
                   age_span = 5,
                   fx_idx = 4)
pop_proj[ , c(1, 2, ncol(pop_proj))]

```


### TMB 

Calculate a population projection in TMB. Carry forward 2000 values for two further periods
to explore projections.

```{r tmb_example}

basepop_init <- as.numeric(burkina.faso.females$baseline.pop.counts)

sx_init <- burkina.faso.females$survival.proportions
sx_init <- cbind(sx_init, `2005` = sx_init[ , "2000"], `2010` = sx_init[ , "2000"])

fx_init <- burkina.faso.females$fertility.rates[4:10, ]
fx_init <- cbind(fx_init, `2005` = fx_init[ , "2000"], `2010` = fx_init[ , "2000"])

gx_init <- burkina.faso.females$migration.proportions
gx_init <- cbind(gx_init, `2005` = gx_init[ , "2000"], `2010` = gx_init[ , "2000"])

log_basepop_mean <- as.vector(log(basepop_init))
logit_sx_mean <- as.vector(qlogis(sx_init))
log_fx_mean <- as.vector(log(fx_init))
gx_mean <- as.vector(gx_init)
  
data <- list(log_basepop_mean = log_basepop_mean,
             logit_sx_mean = logit_sx_mean,
             log_fx_mean = log_fx_mean,
             gx_mean = gx_mean,
             srb = rep(1.05, ncol(sx_init)),
             age_span = 5,
             n_steps = ncol(sx_init),
             fx_idx = 4L,
             fx_span = 7L,
             census_log_pop = log(burkina.faso.females$census.pop.counts),
             census_year_idx = c(4L, 6L, 8L, 10L))
par <- list(log_tau2_logpop = 0,
            log_tau2_sx = 0,
            log_tau2_fx = 0,
            log_tau2_gx = 0,
            log_basepop = log_basepop_mean,
            logit_sx = logit_sx_mean,
            log_fx = log_fx_mean,
            gx = gx_mean)

obj <- make_tmb_obj(data, par)

obj$fn()

matrix(obj$report()$population, nrow = 17)[ , data$census_year_idx]

input <- list(data = data, par_init = par)
fit <- fit_tmb(input)

fit[1:6]

```

Sample from posterior distribution and generate outputs

```{r sample_tmb}
fit <- sample_tmb(fit)

colnames(fit$sample$population) <- 1:ncol(fit$sample$population)

init_pop_mat <- ccmppR(basepop_init, sx_init, fx_init, gx_init,
                       srb = rep(1.05, ncol(sx_init)), age_span = 5, fx_idx = 4)

df <- crossing(year = seq(1960, 2015, 5),
               sex = "female",
               age_group = c(sprintf("%02d-%02d", 0:15*5, 0:15*5+4), "80+")) %>%
  mutate(init_pop = as.vector(init_pop_mat))

census_pop <- crossing(sex = "female",
                       age_group = c(sprintf("%02d-%02d", 0:15*5, 0:15*5+4), "80+")) %>%
  bind_cols(as.tibble(burkina.faso.females$census.pop.counts)) %>%
  gather(year, census_pop, `1975`:`2005`) %>%
  type_convert(cols(year = col_double()))

df <- df %>%
  left_join(census_pop) %>%
  bind_cols(as.tibble(fit$sample$population)) %>%
  gather(sample, value, `1`:last_col()) %>%
  group_by(year, sex, age_group)

agepop <- df %>%
  group_by(year, sex, age_group) %>%
  summarise(init_pop = mean(init_pop),
            census_pop = mean(census_pop),
            mean = mean(value),
            lower = quantile(value, 0.025),
            upper = quantile(value, 0.975))

totalpop <- df %>%
  group_by(year, sample) %>%
  summarise(init_pop = sum(init_pop),
            census_pop = sum(census_pop),
            value = sum(value)) %>%
  group_by(year) %>%
  summarise(init_pop = mean(init_pop),
            census_pop = mean(census_pop),
            mean = mean(value),
            lower = quantile(value, 0.025),
            upper = quantile(value, 0.975))
```

```{r, fig.height = 3, fig.width = 4, fig.align = "center"}
ggplot(totalpop, aes(year, mean, ymin = lower, ymax = upper)) +
  geom_ribbon(alpha = 0.2) +
  geom_line() +
  geom_line(aes(y = init_pop), linetype = "dashed") +
  geom_point(aes(y = census_pop), shape = 4, color = "darkred", stroke = 2) +
  scale_y_continuous("Total population (millions)", labels = scales::number_format(scale = 1e-6)) +
  expand_limits(y = 0) +
  labs(x = NULL) +
  theme_light() +
  ggtitle("BFA females: total population")
```

```{r, fig.height = 5, fig.width = 7, fig.align = "center"}
ggplot(agepop, aes(age_group, mean, ymin = lower, ymax = upper, group = 1)) +
  geom_ribbon(alpha = 0.2) +
  geom_line() +
  geom_line(aes(y = init_pop), linetype = "dashed") +
  geom_point(aes(y = census_pop), color = "darkred") +
  facet_wrap(~year, scales = "free_y") + 
  scale_y_continuous("Total population (thousands)", labels = scales::number_format(scale = 1e-3)) +
  expand_limits(y = 0) +
  theme_light() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid = element_blank()) +
  ggtitle("BFA females: population by age")
```



## Development notes

The simulation model is implemented as templated C++ code in `src/ccmpp.h` 
such that the simulation model may be developed as a standalone C++ library
that can be called by other software without requiring R-specific code features.

* Presently the CCMPP model is implemented as a sparse Leslie matrix formulation. 
  In future, this will probably be split into multiple steps to save intermediate
  computations of demographic events (births, deaths, migrations).
* In future a class structure will probably be developed.
* Specifying static dimensions for the state space may improve efficiency. This 
  should be possible for common options (5x5 year, 1x1 year) through templating.
  
The file `src/ccmppR.cpp` contains R wrapper functions for the model simulation 
via [Rcpp](http://dirk.eddelbuettel.com/code/rcpp.html) and 
[RcppEigen](http://dirk.eddelbuettel.com/code/rcpp.eigen.html).
  
The objective function (the negative log posterior) is implemented in templated 
C++ code using probability functions from the Template Model Builder 
([TMB](https://github.com/kaskr/adcomp/wiki)) R package. Implementation in TMB
provides gradient functions via automatic differentiation (AD).

* Further investigation is needed about the portability of AD objective function
  DLLs outside of the R environment.

TMB also contains R functions for approximation integration of arbitrary random
effects via Laplace approximation.

TMB model code and testing are implemented following templates from the 
[`TMBtools`](https://github.com/mlysy/TMBtools) package with guidance for package 
development with both TMB models and Rcpp code. 
  * To add a new TMB model, save the model template in the `src/TMB` with extension `.pp`. The model name must match the file name. The signature is slightly different -- see other `.hpp` files for example.
  * Call `TMBtools::export_models()` to export the new TMB model to the meta-model list.
  * When constructing the objective function with `TMB::MakeADFun`, use `DLL= "leapfrog_TMBExports"` and add an additional value `model = "<model_name>"` to the `data` list.
